Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-01-06 supplement.tex
reinference from simulated data

While simulations of datasets of this size are rapid, computing LD statistics
is computationally burdensome. We simulated 500 1Mb regions under our best fit
single origin, continuous migration, and merger with stem migration models.
Using these simulated regions, we constructed 500 bootstrap replicate datasets
by sampling with replacement. Each of these 500 replicate sets were fit using
the models reported in the main text, and we compared log-likelihoods across
each fit (Figure~\ref{fig:refit-lls}). In each case, even in the simpler single
origin simulations, the correct model was recovered.  We conclude that our
inferences are not simply fitting statistical noise, and the observed
differences in log-likelihoods indicate a better recapitulation of the true
history.

\begin{table}[ht]
    \caption{
        \label{tab:confusion-matrix}
        \textbf{Something like a confusion matrix}
        \aprcomment{fix/fill caption}
    }
    \centering
    \begin{tabular}[t]{lcccc}
        \toprule
        Simulated model & Single origin & Cont. migration & Merger w/out stem mig. & Merger w/stem mig. \\
        \midrule
        Single origin & 494 & 3 & 1 & 2 \\
        Cont. migration & 0 & 500 & 0 & 0 \\
        Merger w/stem mig. & 0 & 0 & 1 & 499 \\
        \bottomrule
    \end{tabular}
\end{table}

The single origin model is nested within the more complex models (with
migration from the additional stem population fixed to zero), those more
complex models can always recover the log-likelihood found by the single origin
model. We therefore explored the broader parameter space that included gene
flow from the additional stem population. In these fits, the optimization
converged either back to the nested single origin model, or it found a local
optimum with lower log-likelihood than the simpler model (see first panel in
Figure~\ref{fig:refit-lls}).

\begin{figure}[htb]
    \centering
    \includegraphics{./figures/supp-refit-lls.pdf}
    \caption{
        \textbf{Fits of four models to simulated data.} In three scenarios, a
        simple single origin model and our preferred continuous migration and
        merger with stem migration models, we simulated 500 replicates of 500
        Mb with the same populations and sample sizes from our reported best
        fit models. We then fit the four models reported in the main text to
        each of these simulated datasets and compared log-likelihoods. In each
        case, the simulated model is correctly chosen, even under the single
        origin scenario. We conclude our more complex models (with more
        parameters) are not simply fitting statistical noise.
    }
    \label{fig:refit-lls}
\end{figure}

>>>>

While simulations of datasets of this size are rapid, computing LD statistics
is computationally burdensome. We simulated 500 1Mb regions under our best fit
single origin, continuous migration, and merger with stem migration models.
Using these simulated regions, we constructed 500 bootstrap replicate datasets
by sampling with replacement. Each of these 500 replicate sets were fit using
the models reported in the main text, and we compared log-likelihoods across
each fit (Figure~\ref{fig:refit-lls}). The correct model was recovered in the
vast majority of cases based on log-likelihoods alone
(Table~\ref{tab:confusion-matrix}). We conclude that our inferences are not
simply fitting statistical noise, and the observed differences in
log-likelihoods indicate a better recapitulation of the true history.

\begin{table}[ht]
    \caption{
        \label{tab:confusion-matrix}
        \textbf{Model choice from fits to simulated data.} We simulated three
        models, based on the best fit parameters from fits to data. Using these
        simulated datasets, we reinferred four models that we tested in our
        analyses in the main text, including simple (single origin) and complex
        models. Among 500 fits, we almost exclusively recover the simulated
        model, even when fitting more complex (i.e., more heavily
        parameterized) models to simpler simulated models (such as the single
        origin model).
    }
    \centering
    \begin{tabular}[t]{lcccc}
        \toprule
        Simulated model & Single origin & Cont. migration & Merger w/out stem mig. & Merger w/stem mig. \\
        \midrule
        Single origin & 500 & 0 & 0 & 0 \\
        Cont. migration & 0 & 500 & 0 & 0 \\
        Merger w/stem mig. & 0 & 0 & 10 & 490 \\
        \bottomrule
    \end{tabular}
\end{table}

The single origin model is nested within the more complex models (with
migration from the additional stem population fixed to zero), those more
complex models can always recover the log-likelihood found by the single origin
model. We therefore explored the broader parameter space that included gene
flow from the additional stem population. In these fits, the optimization
converged either back to the nested single origin model (in which case the
single origin model was chosen), or it found a local optimum with lower
log-likelihood than the simpler model (see first panel in
Figure~\ref{fig:refit-lls}).

\begin{figure}[htb]
    \centering
    \includegraphics{./figures/supp-refit-lls.pdf}
    \caption{
        \textbf{Log-likelihoods of fits of four models to simulated data.} In
        three scenarios, a simple single origin model and our preferred
        continuous migration and merger with stem migration models, we
        simulated 500 replicates of 500 Mb with the same populations and sample
        sizes from our reported best fit models. We then fit the four models
        reported in the main text to each of these simulated datasets and
        compared log-likelihoods. In each case, the simulated model is
        correctly chosen, even under the single origin scenario. We conclude
        our more complex models (with more parameters) are not simply fitting
        statistical noise.
    }
    \label{fig:refit-lls}
\end{figure}
